---
title: "Gov 50 Final Project"
author: "Abia S. Khan"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

Recently, school boards across the country have been grappling with a number of political challenges and have often been weaponized to advance certain political agendas. From issues of public health (COVID and masking) to LGBTQ+ issues (North Carolina bathroom bills) to the curriculum (Critical Race Theory, book bans), school boards have had to address political issues that aren’t explicitly related to education. While all these issues remain pertinent for school boards across the country, the central mission of a school board remains unchanged: ensuring high student achievement, especially with concerns about learning loss following the pandemic. At the end of the day, school boards and districts are held accountable to the standards set by the state. Are they continuing to uphold these standards set by the state while addressing these additional political issues, or are these political issues serving as a distraction from addressing student achievement, leading into my main question, What is the effect of political extremism on school boards?

## Methods

In order to determine if there are any patterns in spending by political action committees (PACs) or candidate committees (federal or local), I compiled the campaign finance reports of candidates in school board races over a period of time. However, getting to this point was a multi-step process given the federalist nature of education and the fact that it is ultimately up to the individual states to determine how they govern their schools. As a result, there is significant variation from state to state regarding the governance structure of school districts, the number of school districts, how many layers of bureaucracy there are, etc.

The first step I took in compiling this data was by going to the U.S. Census Bureau and pulling the list of school districts (along with what county in the state they are associated with) they have recorded. The U.S. Census Bureau has collected this data every year up until 2022, so the data I am using for the purposes of this thesis is the 2022 report. This dataset had 18,990 districts across all 50 states and the District of Columbia.

For this project, I analyze districts that represent entire counties (i.e., counties that only have a single school district representing the region). These districts will generally be bigger in size and represent more stakeholders than in districts where they are one of many in a single county. As a result, these will likely be higher-profile races, attracting more money and increasing the incentives for groups with a vested interest in the district to recruit candidates for a seat on the board or direct financial resources toward candidates who align with their political views. With this, patterns of spending will be easier to track than in smaller districts where the school board race is a lower-profile race. Furthermore, campaign finance laws for school board races dictate that campaign finance reports be filed with the county that the district is situated in (with a few exceptions); Therefore, stratifying my analysis by school districts that represent entire counties will allow for easier data collection.

After cleaning up the data based on these conditions, I was left with 953 districts. However, I noticed that in some states, single districts were situated within multiple counties (and were often state departments of education and therefore were not governed by individual school boards but the state itself, which was the case in states like Hawaii and New York) or some districts were under mayoral control (such as the District of Columbia). After removing these districts, I was left with 942 districts representing 37 states. (For the purposes of this project, I focused on five school districts in Florida.)

From there, I pulled the campaign finance reports from each county’s school board race from 2012-2022 from their respective campaign finance databases (often housed in the websites of local boards of elections, county recorders office, or the offices of the secretaries of state). I organized the campaign finance reports with the following columns:

Candidate First Name
Candidate Last Name
State
County
District/Region
Year of Race
Total Contributions
Total PAC Contributions
Success of Race (with 1 meaning the candidate was elected, 0 otherwise)

The independent variable I analyze is campaign contributions by national donors and/or political action committees. More specifically, I look at campaign contributions given by those with a certain political slant, such as unions, parent advocacy groups, school choice advocacy groups, etc. I use these groups as a mechanism for measuring the presence of political polarization within school boards, where increased spending in a particular race indicates an attempt for a certain political ideology or agenda to be pushed in via a particular candidate. Furthermore, I track these patterns over time in order to see if there are any changes in spending at a particular point in time.

The dependent variables I analyze in my quantitative analysis are the volume of campaign contributions from political action committees (PACs) or candidate committees (federal or local) to school board candidates. The proportion of PAC and candidate committee contributions (relative to their total contributions) can show the level of outside interest within the race, going against the typical hyper-localized aspect of school board races.

```{r}
library(tidyverse)
campaign_finance <- read.csv("data/campaign_finance.csv")
campaign_finance$Year <- as.character(campaign_finance$Year)

avg_finance <- campaign_finance |>
  group_by(Year) |>
  summarize(mean(Total.Raised), mean(PAC.Contributions))

avg_pac_cont <- avg_finance |>
  ggplot(mapping = aes(x = Year, y = `mean(PAC.Contributions)`)) +
  geom_col() +
  labs(x = "Year", y = "Average PAC Contribution (in thousands)")

avg_pac_cont
```

## Results


## PAC Contributions as a Proportion of Total Campaign Contributions
```{r}
campaign_finance <- campaign_finance |>
  mutate(pac_prop = (PAC.Contributions/Total.Raised) * 100)
```

```{r}
plot_1 <- campaign_finance |>
  ggplot(mapping = aes(x = pac_prop)) +
  geom_histogram() +
  labs(x = "PAC Contributions as a Proportion of Total Campaign Contributions", y = "Frequency", Title = "Frequency of the Proportion of PAC Contributions")
```

```{r}
plot_1
```

This figure shows the occurrence of PAC and candidate committee contributions as a proportion of total contributions (in percentage) received by school board candidates. This figure reaffirms the findings in the previous plot, showing a concentration of the proportion of PAC and candidate committee contributions around the 0-15% mark.

However, there are isolated instances of school board candidates receiving over 40% of their total contributions from PACs or candidate committees, with one candidate receiving nearly all of their contributions from a PAC or candidate committee. Given that PACs and candidate committees seldom get involved in a school board race at all, it is highly unusual that they make up a significant proportion of a school board candidate’s financial resources, and in one instance, effectively funding the entirety of a school board candidate’s race. This is another signal that outside groups are starting to look to school boards as an institution to push their political agenda through.

## Total contributions vs. PAC Contributions
```{r}
campaign_finance <- campaign_finance |>
  mutate(raised = Total.Raised/1000,
         PAC = PAC.Contributions/1000)

plot_2 <- campaign_finance |>
  ggplot(mapping = aes(x = raised, y = PAC)) +
  geom_point() +
  labs(x = "Total Contributions (in thousands)", y = "Total PAC Contributions (in thousands")
```

```{r}
plot_2
```

This figure shows the relationship between the total contributions that a school board candidate receives and the total PAC and candidate committee contributions they receive. This plot shows that most of the data points are clustered between $0-$50,000 on the x-axis and $0-$10,000 on the y-axis. This is expected given that school board races are at the bottom of the ballot, meaning that they end up being low-information and low-turnout races, and therefore low-budget races. Adding onto the fact that school board races are often not competitive, school board candidates do not have to raise much money in order to win a race (relative to other local and federal races) and this figure affirms that to a certain extent. Furthermore, PACs and candidate committees typically do not get involved in school board races, whether that be through an endorsement, candidate recruitment, etc. Therefore, school board candidates typically receive very few contributions from PACs and candidate committees, with perhaps modest contributions coming from local labor unions.

However, it is clear that there are some points in the dataset that are further away from the cluster, indicating increased fundraising by the school board candidate themselves as well as increased contributions by PACs and candidate committees to school board candidates. While the former can be attributed to the rising costs of campaigns over time given their lengthened timeline, the latter is more difficult to explain, especially when the hyper-local nature of school board races has not changed. While the cost of political campaigns in general may be increasing, that does not explain why PACs and candidate committees would get involved financially in a school board races, especially when at the end of the day, financial resources are still scarce and candidates at the top of the ballot are in the greatest need of those resources.

These outliers in PAC and candidate committee contributions indicate that there may be some groups with a vested interest in either that particular school board or the region the school board serves and is attempting to influence the outcome of the races by pouring financial resources into candidates who align with their political agenda.

## Effect of Contributions on Election Results

```{r}
regression <- lm(Elected ~ Total.Raised + PAC.Contributions, data = campaign_finance)

modelsummary::modelsummary(regression)
```

This regression shows that without any contributions, candidates will be elected 28.4% of the time. The coefficients for the Total.Raised and PAC.Contribution columns are 0, indicating that an increase in either of these values has no effect on a school board candidate's success.

## Conclusion
With these data, there's not much support for my hypothesis. In general, the data shows that contributions have no effect on election success, and that spending patterns (across both PAC contributions and other contributions) are rather stagnant and have no particular rhyme or rhythm.

A limitation of the data is that school board races are typically low-dollar races, and any changes in contributions have happened very recently, meaning that there's only 1-2 cycles worth of data to draw from. Furthermore, any increases in spending in any particular race are likely to be outliers in the general dataset, making the data noisy.

A few things could be done in order to increase the quality of my analysis. If I had more time, I would include more data points from different states and try to have as many districts as possible represented in my sample. Furthermore, while there might not be a systemic shift in PAC spending in school board races, the outliers in the dataset might also be telling (in terms of which districts are being targeted and by who), so doing a qualitiative analysis of these outliers could give additional information about the nature of school board races.