---
title: "Gov 50 Final Project"
author: "Abia S. Khan"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to campaign finance reports in school board races. I'd like to find patterns in outside spending in these local races.

## Project Proposal

What is the effect of political polarization on American school boards? I hypothesize that in recent years, there has been more attention on school boards due to the ways that national issues have intersected with this local institution. While school boards are supposed to address issues that are locally relevant, they also at times have to address issues that are more nationally focused, such as public health measures during the COVID-19 pandemic, critical race theory, LGBTQ+ issues, etc. Since school boards are having to take a position on these issues and implement policies over them, they have increased on visibility and as a result become the target of national groups. This can come in many forms, from the mobilization of people to recruiting candidates to run in competitive school board races to pouring money into school board races. In order to explore this topic, I will analyze campaign finance data from a number of school board races over time and focus on campaign contributions from political action committees as well as national donors (which is my explanatory variable). My outcome variable is the result of these races. Patterns of increased spending in school board races that lead to outcomes that align with these donor networks will indicate that political polarization is seeping into school board races in a tangible way.

## Data Visualization
```{r}
library(tidyverse)
campaign_finance <- read.csv("data/campaign_finance.csv")
```


## PAC Contributions as a Proportion of Total Campaign Contributions
```{r}
campaign_finance <- campaign_finance |>
  mutate(pac_prop = (PAC.Contributions/Total.Raised) * 100)
```

```{r}
plot_1 <- campaign_finance |>
  ggplot(mapping = aes(x = pac_prop)) +
  geom_histogram() +
  labs(x = "PAC Contributions as a Proportion of Total Campaign Contributions", y = "Frequency", Title = "Frequency of the Proportion of PAC Contributions")
```

```{r}
plot_1
```

This figure shows the occurrence of PAC and candidate committee contributions as a proportion of total contributions (in percentage) received by school board candidates. This figure reaffirms the findings in the previous plot, showing a concentration of the proportion of PAC and candidate committee contributions around the 0-15% mark.
However, there are isolated instances of school board candidates receiving over 40% of their total contributions from PACs or candidate committees, with one candidate receiving nearly all of their contributions from a PAC or candidate committee. Given that PACs and candidate committees seldom get involved in a school board race at all, it is highly unusual that they make up a significant proportion of a school board candidate’s financial resources, and in one instance, effectively funding the entirety of a school board candidate’s race. This is another signal that outside groups are starting to look to school boards as an institution to push their political agenda through.

## Total contributions vs. PAC Contributions
```{r}
campaign_finance <- campaign_finance |>
  mutate(raised = Total.Raised/1000,
         PAC = PAC.Contributions/1000)

plot_2 <- campaign_finance |>
  ggplot(mapping = aes(x = raised, y = PAC)) +
  geom_point() +
  labs(x = "Total Contributions (in thousands)", y = "Total PAC Contributions (in thousands")
```

```{r}
plot_2
```

This figure shows the relationship between the total contributions that a school board candidate receives and the total PAC and candidate committee contributions they receive. This plot shows that most of the data points are clustered between $0-$50,000 on the x-axis and $0-$10,000 on the y-axis. This is expected given that school board races are at the bottom of the ballot, meaning that they end up being low-information and low-turnout races, and therefore low-budget races. Adding onto the fact that school board races are often not competitive, school board candidates do not have to raise much money in order to win a race (relative to other local and federal races) and this figure affirms that to a certain extent. Furthermore, PACs and candidate committees typically do not get involved in school board races, whether that be through an endorsement, candidate recruitment, etc. Therefore, school board candidates typically receive very few contributions from PACs and candidate committees, with perhaps modest contributions coming from local labor unions.
However, it is clear that there are some points in the dataset that are further away from the cluster, indicating increased fundraising by the school board candidate themselves as well as increased contributions by PACs and candidate committees to school board candidates. While the former can be attributed to the rising costs of campaigns over time given their lengthened timeline, the latter is more difficult to explain, especially when the hyper-local nature of school board races has not changed. While the cost of political campaigns in general may be increasing, that does not explain why PACs and candidate committees would get involved financially in a school board races, especially when at the end of the day, financial resources are still scarce and candidates at the top of the ballot are in the greatest need of those resources.
These outliers in PAC and candidate committee contributions indicate that there may be some groups with a vested interest in either that particular school board or the region the school board serves and is attempting to influence the outcome of the races by pouring financial resources into candidates who align with their political agenda.